Empirical Bayes analysis of \RNAseq{} data for detection of gene expression heterosis

### 2015-09-01 response.tex

\response{We thank the AE and two reviewers for their comments which dramatically improved the manuscript.}

Associate Editor comments:

\comment{The proposed model appears to be an extension of Ji et al. (2014) albeit with a differently specified likelihood function (negative binomial or Poisson-gamma) as perceived to be more appropriate for count data relative to the Gaussian likelihood function pursued in Ji et al. (2014). Both reviewers indicate that they are little underwhelmed by the performance of your proposed procedure relative to some of the existing methods that you compare against (i.e., ShrinkBayes, edgeR and baySeq).  Furthermore, they seem somewhat concerned about an ‘apples vs. oranges’ types of comparisons since the proposed hypothesis testing strategy in the proposed method is really not an option in the other methods.}

\response{We have assessed the Ji et al. approach and included it in Sections 3.2 and 3.3. Although the Ji et al.\ approach performed better than edgeR and baySeq, its performance was considerably worse than the performance of the methods we proposed.  Regarding the `apples vs.\ oranges comparison', we agree that the proposed hypothesis testing strategy is not available in other methods.  This was part of the motivation for developing our methods.  Our simulations show that doing the best we know how to do with existing methods is worse than doing what we proposed.  We think the comparison is relevant and answers a question that some readers would have raised had we not included methods like edgeR and baySeq in our simulation.}

@@

\comment{Perhaps a reference to the promise of allele specific expression analyses (Discussion section) for inferring upon gene-expression heterosis might be useful to point readers into other directions with this work (Bell et al., 2013; Wei and Wang, 2013).}

\response{Thanks for pointing out these papers.  We have added a new concluding paragraph to the paper that cites the papers you mentioned as well as two others.}

@@

\comment{I realize that Ji et al.’s (2014) work was intended for microarray data, or more specifically Gaussian likelihoods, but there has been some work that suggests that Gaussian likelihood specifications in MAANOVA may, oddly enough, have better properties for the analysis of RNA-seq data than more popular methods specifically intended for count data (edgeR and DESeq).  See Reeb and Steibel (2013) and also comments from one reviewer along these lines.  Since Ji et al.’s work also derives from ISU, have you assessed its performance relative to your proposed method on RNA-seq data?}

\response{We have included Ji et al.'s work as another alternative method, see Sections 3.2 and 3.3.}

\comment{One seemingly glaring omission from this work is any comparative assessment of the proposed versus other methods (e.g. ShrinkBayes) on the actual maize dataset in Section 4.  Why is that missing? The comments made by one reviewer regarding a relative assessment of computing time relative to the other methods deserves attention.}

\response{We have added a comparison of the computing time required for each method in Section 4 of the revised manuscript.}

@@

\comment{Page 2 line 35.  “We CONSIDER an…”}

\response{We fixed the error.  Thanks for catching this.}}

\comment{Page 3 Line 7.  I don’t it is sufficient to mention the statistical software “Stan”.  Please provide a suitable description and citation.}

\response{We agree and have added the appropriate citation.}}

@@

\comment{Page 3 Line 26.  There are several ways to construct a Poisson-gamma distribution, all leading to different specifications of the relationship between the mean and variance of a negative binomial.  I apologize for the admittedly vain self-citation but there is one specification, for example, that specifies the variance to be a linear function of the mean of a negative binomial…see Equation 2.12b in  Tempelman and Gianola (1996).  I guess I’m not sure why most RNA-seq data analysts gravitate to a specification where the variance is specified to be linear and quadratic function of the mean as you do here.}

\response{You have raised in interesting point.  If $y_i|\lambda_i$ is Poisson with mean $\lambda_i$, and $\lambda_i$ is ${\rm gamma}(\alpha_i,\beta)$ for all $i=1,\ldots, n$, then (for all $i$) the marginal distribution of $y_i$ is negative binomial with mean $\mu_i$ and variance $\phi\mu_i$, where $\mu_i=\alpha_i\beta$ and $\phi=1+\beta$.  Thus, if we assume the $\beta$ parameter is constant across observations, we have a linear mean-variance relationship.  On the other hand, if $y_i|\lambda_i$ is Poisson with mean $\lambda_i$, and $\lambda_i$ is ${\rm gamma}(\alpha,\beta_i)$ for all $i=1,\ldots, n$, then (for all $i$) the marginal distribution of $y_i$ is negative binomial with mean $\mu_i$ and variance $\mu_i+\phi\mu_i^2$, where $\mu_i=\alpha\beta_i$ and $\phi=1/\alpha$.  Thus, if we assume the $\alpha$ parameter is constant across observations, we have a quadratic mean-variance relationship.  Past research suggests that RNA-seq data that includes biological (rather than only technical) variation tends to exhibit a quadratic mean-variance relationship.  Thus, we have focused on the latter type of negative binomial model specification.  It is conceivable that the linear mean-variance relationship could be appropriate for some subset of genes in an RNA-seq dataset.  That is an interesting model selection question, but one that we think should be left for future work focused specifically on this issue. }

@@

\comment{Page 3. Bottom.  Shouldn’t you at least provide a justification when one would use a normal prior and when one would or should use a Laplace prior?}

\response{We have assessed the use of normal distributions see Section 3.3 and also provided rationale for why Laplace distributions would be expected to perform better in this scenario at the end of Section 2.1. Our simulation results in Figure 2 show that our Laplace model specification is preferred over the normal model specification.}

@@

\comment{Page 4, bottom half.  The strategy described here, also somewhat attested to by a reviewer, is a little disconcerting with respect to the “double estimation” of gene-specific parameters…first to estimate they hyperparameters and then again conditioned on the estimates of these hyperparameters.  Perhaps you should characterize that these second-stage gene-specific estimates are shrunk relative to those estimates used to estimate the hyperparameters and that they then have lead to predictions with better properties?  Ideally, I would have liked to see these hyperparameters be also formally estimated together with the gene specific parameters within one MCMC analyses…authors should address why they didn’t do this.}

\response{We have added a paragraph to the beginning of Section 2.2 to share our experience with a fully MCMC analysis which took too long and had poor mixing. We have added a new Figure 3, which shows that the second-stage gene-specific estimates are shrunk relative to the original edgeR estimates.}

@@

\comment{Page 5, lines 37-47.  Since you’re doing MCMC, why not simply tabulate the number of times samples of $\mu_3$ exceeds either $\mu_1$ or $\mu_2$….and/or the number of times that $\mu_3$ is less than $\mu_1$ or $\mu_2$.  I suspect that what your test is really doing….in which case it might be useful to explicitly state this equivalence as such.}

\response{Yes, that is what the test is doing. We have now stated the case for LPH, i.e. ``$\delta_g < -|\alpha_g|$ (equivalently $\mu_{g3} < \min\{\mu_{g1},\mu_{g2}\}$).''

\comment{Page 6, top.  Is it possible then to assess your proposed method using Laplace versus normal priors to assess if that seems to be contributing to any differences between ShrinkBayes and your proposed method?  Maybe that is the real reason for the small improvements?...., and not the differences in sophistication of the constructed statistical tests?}

\response{We have now assessed the use of normal distributions. The results in Section 3.3 and Figure 2 show that there is an advantage to using Laplace distributions instead of normals.  Our eBayes implementation of our model performs quite similarly to our ShrinkBayes implementation of our model.}

@@

\comment{Page 7, line 15.  Why delete the low counts?  Data quality issues?}

\response{ShrinkBayes failed for us when too many zeros were present. We do not believe the comparison of other methods would change if we did not remove genes with low counts.  In practice, it is common to remove low count genes from RNA-seq datasets prior to analysis because these genes contain little information about differential expression and lead to computational difficulties for some methods. With that being said, we did not remove any genes from the real data analysis.}

@@

\comment{Page 7, lines 17-22.  Can you provide us any information on how well the hyperparameters were estimated based on your empirical Bayes strategy?}

\response{Since we did not simulate data from our model, there are no true values of the hyperparameters and therefore we cannot directly compare our estimates to a truth.  We did try to point out in the Discussion that we believe these estimates could be improved.}

\comment{Page 8, line 37.  It is not obvious to me why $H_g2*$ (are the two parents the same?) is a necessary component of a heterosis (overdominance) test!?}

\response{It is not necessary, but it is sufficient. If parental expressions are the same and the hybrid is not, then the hybrid must be exhibiting either LPH or HPH.}

@@

\comment{Page 9, lines 23-25.  In tandem with another reviewer’s comments, the real question is why?  Is it because of the suboptimality of the test that you contrived for ShrinkBayesA?}

\response{In Section 3.3, we highlighted the differences between ShrinkBayes and the eBayes approaches, but do not try to specifically state which difference is the cause.}

\comment{Page 10, Line 29.  Don’t you mean “nadir” rather than peak?}

\response{No, although we do mean ridge rather than peak, i.e. there are many genes with effect size of zero and probabilities below 0.5.}

### 2015-09-01 response.tex

\comment{Page 2 line 35.  “We CONSIDER an…”}

\response{We fixed the error.  Thanks for catching this.}

\comment{Page 3 Line 7.  I don’t it is sufficient to mention the statistical software “Stan”.  Please provide a suitable description and citation.}

\response{We agree and have added the appropriate citation.}

@@

\comment{Page 4, bottom half.  The strategy described here, also somewhat attested to by a reviewer, is a little disconcerting with respect to the “double estimation” of gene-specific parameters…first to estimate they hyperparameters and then again conditioned on the estimates of these hyperparameters.  Perhaps you should characterize that these second-stage gene-specific estimates are shrunk relative to those estimates used to estimate the hyperparameters and that they then have lead to predictions with better properties?  Ideally, I would have liked to see these hyperparameters be also formally estimated together with the gene specific parameters within one MCMC analyses…authors should address why they didn’t do this.}

\response{We have added a paragraph to the beginning of Section 2.2 to share our experience with a fully MCMC analysis which took too long and had poor mixing. We have added a new Figure 3, which shows that the second-stage gene-specific estimates are shrunk relative to the original edgeR estimates.}

\comment{Page 5, lines 37-47.  Since you’re doing MCMC, why not simply tabulate the number of times samples of $\mu_3$ exceeds either $\mu_1$ or $\mu_2$….and/or the number of times that $\mu_3$ is less than $\mu_1$ or $\mu_2$.  I suspect that what your test is really doing….in which case it might be useful to explicitly state this equivalence as such.}

\response{Yes, that is what the test is doing. We have now stated the equivalence for both LPH and HPH, e.g. for LPH: $\delta_g < -|\alpha_g|$ is equivalent to $\mu_{g3} < \min(\mu_{g1},\mu_{g2})$.}

@@

\comment{Page 7, lines 17-22.  Can you provide us any information on how well the hyperparameters were estimated based on your empirical Bayes strategy?}

\response{Since we did not simulate data from our model, there are no true values of the hyperparameters and therefore we cannot directly compare our estimates to a truth.  We did try to point out in the Discussion that we believe these estimates could be improved.}

\comment{Page 8, line 37.  It is not obvious to me why $H_g2*$ (are the two parents the same?) is a necessary component of a heterosis (overdominance) test!?}

\response{Yes, that is the hypothesis that the two parents are the same, but the hybrid is different. It is not necessary, but it is sufficient. If parental expressions are the same and the hybrid is not, then the hybrid must be exhibiting either LPH or HPH.}

@@

\response{The area under the curve attempts to do exactly this, i.e. a single metric to compare the methods. We do not try to over-emphasize the difference between the ShrinkBayes and eBayes methods, which are both based on our model, as we do not believe there is much of a difference.}

Reviewer \#4

\comment{1. The authors propose a method to assess "gene expression heterosis."   While it is mentioned that this type of heterosis is a possible explanation for phenotypic heterosis, I'd like more discussion regarding this point.  The two references seem a bit outdated (for bioinformatics) so I'm wondering if this explanation is still being considered.  If it is, than I'd like to see more discussion about how the analytic results would be used to investigate this explanation.   Wouldn't scientists have specific phenotypic traits in mind and know of genes associated with these traits?  I'm a little confused why the search through the whole genome for this type of heterosis if they're looking to draw inference on the association.}

\response{We now include a reference to Chen (2013), a Nature Reviews Genetics article entitled ``Genomic and epigenetic insights into the molecular bases of heterosis'' as well as three other articles published in 2012 or later. Even though scientists may have specific candidates. An exploratory analysis to suggest other genes seems useful.}

@@

\response{We simulate data from the negative-binomial portion of the model, but do not simulate from the hierarchical portion of the model. Instead, we use edgeR estimates for gene-specific parameters which provide some shrinkage for the overdispersion parameters but none for the other gene-specific parameters. We have tried to make this clear in Section 3.1. In addition, we have performed the analysis by simulating entirely from our model and achieve similar results.}

\response{We also tried to emphasize in Section 2 that ``our'' approach includes the implementation in ShrinkBayes as the model itself is novel and eBayes and ShrinkBayes are just two different approaches to estimation.}

\comment{a. How robust do you feel your method is?  There is no discussion along these lines.}

\response{In the discussion, we mention the departures from model assumptions observed. Despite these violations, the results seems reasonable and thus suggest some level of robustness. We believe that further analysis of robustness is better left to future work specifically devoted to addressing this question.}

### 2015-09-02 ebayes_rnaseq_heterosis.tex
small typos

To use \RNAseq{} counts to identify genes displaying expression heterosis, we build a hierarchical model to borrow information across gene-variety means and across gene-specific overdispersion parameters, estimate the parameters using an empirical Bayes procedure, and calculate empirical Bayes posterior probabilities for extreme heterosis.

@@

In the maize context that motivates this work, we are interested in extreme heterosis (EH), either low-parent heterosis (LPH) or high-parent heterosis (HPH), in gene expression. 
For a specific gene $g$, LPH occurs when average expression in the hybrid is less than the average expression of either parent, i.e. $\delta_g < -|\alpha_g|$ or, equivalently, $\mu_{g3} < \min\{\mu_{g1},\mu_{g2}\}$, and HPH occurs when average expression in the hybrid is greater than the average expression of either parent, i.e. $\delta_g > |\alpha_g|$ or, equivalently, $\mu_{g3} > \max\{\mu_{g1},\mu_{g2}\}$. 
% and high-parent  heterosis (HPH) of gene expression occurs when $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$. For simplicity, we define \emph{heterosis} of gene expression to be the event of LPH or HPH. The most important hypotheses are
% \begin{align}
% H_{g0}:&\mu_{g1}\le \mu_{g3}\le \mu_{g2} \mbox{ or } \mu_{g1}\ge \mu_{g3}\ge \mu_{g2} \mbox{\ \ vs.\ \ } \nonumber \\
% H_{g1}:&\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3},
% \label{e:hypotheses}
% \end{align}
% where $H_{g0}$ indicates no heterosis for gene $g$ and $H_{g1}$ indicates  heterosis for gene $g$. 
% We evaluate these hypotheses based on empirical Bayes estimates of their posterior probabilities, e.g., 
% \begin{align}
% P\left(H_{g1}|y, \hat{\pi}_{MAP}\right) &= P\left(\left.\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3}\right| y, \hat{\pi}_{MAP}\right) \nonumber \\
% &\approx \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\mu_{g3}^{(m)}< \min\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) \mbox{ or } \max\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) < \mu_{g3}^{(m)}\right), \label{e:probs}
% \end{align}
% where $\mu_g^{(m)} = \left(\mu_{g1}^{(m)},\mu_{g2}^{(m)},\mu_{g3}^{(m)}\right)$ is the $m^{th}$ MCMC sample from the empirical Bayes posterior, and $\mathrm{I}(A)$ is 1 if A is true and 0 otherwise. 
We evaluate these probabilities based on empirical Bayes estimates of their posterior probabilities, e.g., 
\begin{equation}
%P(LPH_g|y) \approx 
P(\mbox{LPH}_g|y,\hat{\pi},\hat{\gamma}) = 
%P\left(\left. \mu_{g3} < \min\{\mu_{g1},\mu_{g2}\}\right|y,\hat{\pi},\hat{\gamma}\right) = 
P\left(\left.\delta_g < -|\alpha_g|\right|y, \hat{\pi},\hat{\gamma}\right) = \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\delta_g^{(m)} < -\left|\alpha_g^{(m)}\right|\right), 
\label{e:probs}
\end{equation}
where $\left(\delta_g^{(m)},\alpha_g^{(m)}\right)$ is the $m^{th}$ MCMC sample from the empirical Bayes posterior, and $\mathrm{I}(A)$ is 1 if A is true and 0 otherwise. HPH probability is defined similarly with the inequality reversed and without the negative sign. We construct a ranked list of genes according to the maximum of the gene's LPH and HPH heterosis probabilities. Geneticists can use this list to prioritize future experiments to understand the molecular genetic mechanisms for heterosis.

@@

Using these parameters and normalization factors, we simulated data according to the negative binomial model in \eqref{e:data} independently for each gene. For each simulation, we analyzed a subset of 25,000 genes selected randomly from genes with simulated counts at least one on average and with at most two zeros for each variety across replications.  We repeated this simulation process 10 times for each of 4, 8, and 16 replicates per variety, reusing normalization factors when necessary.

@@

We compared our method to that of \cite{ji2014estimation}, which assumes normality in the response, by modeling the logarithm of the RNAseq count plus one scaled by the normalization factor, i.e. $\log([Y_{gvi}+1])-\gamma_{vi}$. Use of the normalization factor here provides two advantages: 1) counts are properly adjusted for the thoroughness of the sequencing of the sample and 2) for genes with no count variation within variety (which actually occurs in our maize data set), use of the normalization factors allows the approach of \citeauthor{ji2014estimation} to still execute.

@@

With 4 replicates per variety, there does not appear to be much of a difference between \ShrinkBayes{} and the eBayes approaches, but as the number of replicates increases, the eBayes approaches appear to improve relative to \ShrinkBayes{}. Two difference exist between the \ShrinkBayes{} and eBayes approaches that could explain the difference in AUC: 1) \ShrinkBayes{} utilizes a different empirical Bayes approach for estimating both the hyperparameters and the gene-specific parameters and 2) the measure of EH is slightly different due to INLA not providing a full posterior.

There also appears to be a pattern of the eBayes approach with Laplace distributions for the half-parental differences and hybrid effect systematically performing better than the approach that uses normal distributions. We assume this is because the Laplace distributions are better approximations to the true underlying distribution for these parameters.

@@

We used our method to analyze a maize data set \citep{paschold2012complementation} of \RNAseq{} gene expression in parental lines B73 and Mo17 and the hybrid genotype (B73$\times$Mo17). Each variety had four biological replicates measured with Illumina methodology and equipment. Reads were mapped to the whole reference genome using the short reads aligner, NOVOALIGN. For more specifics, please see \cite{paschold2012complementation}.

We analyzed the data using all the methods compared in the previous section. The computation time on a Dell Precision T3600 with a 3.6GHz Intel Xeon processor was 14 seconds for \edgeR{}, 1.3 minutes for \citeauthor{ji2014estimation}, 10 hours for the eBayes approaches, and 17 hours for \baySeq{}. In the eBayes approach, the vast majority of the time is spent on independent MCMC analysis for each gene. Thus we parallelized this step using {\tt doMC} \citep{revolution2014domc} and {\tt plyr} \citep{hadley2011split}. When parallelized across 5 cores, the eBayes approaches took about 2.5 hours.

For the eBayes approach with the Laplace distributions, we estimated the hyperparameters to be $\hat{\eta}_\phi = 3.44$, $\hat{\sigma}_\phi = 2.74$, $\hat{\eta}_\alpha = -0.05$, $\hat{\sigma}_\alpha=0.36$, $\hat{\eta}_\delta = 0.08$, $\hat{\sigma}_\delta=0.33$, $\hat{\eta}_\psi = -2.35$, and $\hat{\sigma}_\psi=0.60$. We then performed independent MCMC analysis for each gene conditional on these hyperparameters. As with the simulation study, we ran 4 chains simultaneously and doubled the number of MCMC iterations until each gene-specific parameter had an effective sample size above 1,000.

Figure \ref{f:gene_specific_estimates} provides point estimates of the gene-specific parameters from the initial \edgeR{} estimation step and after the eBayes with Laplace distribution procedure described in Section \ref{s:ebayes}. 
\begin{figure}
\centerline{\includegraphics[width=\textwidth]{gene_specific_estimates.pdf}}
\caption{Two-dimensional histogram of point estimates from \edgeR{} and posterior means from eBayes (Laplace) along with the $y=x$ diagonal.}
\label{f:gene_specific_estimates}
\end{figure}
This figure shows shrinkage for large absolute values of $\alpha$ and $\delta$ from the \edgeR{} estimation toward $\hat{\eta}_\alpha$ and $\hat{\eta}_\delta$ which are both approximately zero. The figure also shows decreased values for the overdispersion parameter with larger decreases for high and low values of overdispersion. Finally, very little, if any, shrinkage is observed for $\phi$.

With posteriors for all parameters, we can calculate empirical Bayes posterior probabilities for LPH and HPH. For each gene, the quantity of interest is likely to be the maximum of these two probabilities. For each gene with a high probability of either HPH or LPH, the magnitude of the effect is of interest. We calculate 
\begin{equation}
% \mbox{estimated effect size}_g = \left\{ 
% \begin{array}{ll}
% \hat{\mu}_{g3} - \min(\hat{\mu}_{g1},\hat{\mu}_{g2}) & \mbox{if } \hat{\mu}_{g3} < \min(\hat{\mu}_{g1},\hat{\mu}_{g2}) \\
% \hat{\mu}_{g3} - \max(\hat{\mu}_{g1},\hat{\mu}_{g2}) & \mbox{if } \hat{\mu}_{g3} > \max(\hat{\mu}_{g1},\hat{\mu}_{g2}) \\
% 0 & \mbox{otherwise},
% \end{array} 
% \right. 
\mbox{estimated effect size}_g \equiv \left\{ 
\begin{array}{ll}
\hat{\delta}_g - |\hat{\alpha}_g| & \mbox{if } \hat{\delta}_g > \phantom{-}|\alpha_g| \\
\hat{\delta}_g + |\hat{\alpha}_g| & \mbox{if } \hat{\delta}_g < -|\alpha_g| \\
0 & \mbox{otherwise},
\end{array}.
\right. 
\label{e:effect_size}
\end{equation}
Thus, the estimated effect size is the difference between hybrid mean and the nearest parent with negative values indicating LPH and positive values indicating HPH. If the hybrid mean is estimated to be between the parents, this effect is defined to be zero.

@@

Although our method appears to hold some advantage over existing methods, we believe our approach can be improved by refining the hierarchical model for the gene-specific parameter distribution. Figure \ref{f:scatterplot} shows marginal and bivariate histograms for the empirical Bayes posterior means for the gene-specific parameters. 
\begin{figure}
\centerline{\includegraphics[width=\textwidth]{estimates}}
\caption{Marginal and bivariate histograms of posterior means for gene-specific parameters for the B73 $\times$ Mo17 maize experiment.}
\label{f:scatterplot}
\end{figure}
These figures show markedly departures from marginal model assumptions, e.g. normality assumptions for $\phi_g$ and $\psi_g$, and independence assumptions, e.g. $\phi_g$ versus $\psi_g$ and $\alpha_g$ versus $\delta_g$. The plot of $\phi_g$ versus $\psi_g$ shows a pattern where the mean overdispersion decreases as the mean expression level increases. The plot of $\alpha_g$ versus $\delta_g$ shows a rotated V pattern where $\delta_g$ appears to be equal to $|\alpha_g|$. This pattern is consistent with Mendel's Law of Dominance where the hybrid has mean expression equal to the parent with higher mean expression.

In addition to improving the hierarchical distribution, we believe better estimates of the parameters of this distribution, i.e. the hyperparameters, will also improve detection of gene-expression heterosis. Our current method, based on moment matching of essentially independently estimated gene-specific parameters, provides consistent estimators as the number of replicates per variety increases. But typically the number of replicates per variety is quite small. In our data there are only four replicates per variety, and therefore asymptotic justifications are deficient. There are a variety of hyperparameter estimation approaches to explore, e.g. expectation-maximization algorithms or fully Bayesian approaches.

>>>>

To use \RNAseq{} counts to identify genes displaying expression heterosis, we build a hierarchical model to borrow information across gene-variety means and across gene-specific overdispersion parameters, estimate the hyperparameters using an empirical Bayes procedure, and calculate empirical Bayes posterior probabilities for extreme heterosis.

@@

In the maize context that motivates this work, we are interested in extreme heterosis (EH), i.e. either low-parent heterosis (LPH) or high-parent heterosis (HPH), in gene expression. 
For a specific gene $g$, LPH occurs when expected expression in the hybrid is less than the expected expression of either parent, i.e. $\mu_{g3} < \min\{\mu_{g1},\mu_{g2}\}$ or, equivalently, $\delta_g < -|\alpha_g|$, and HPH occurs when expected expression in the hybrid is greater than the exptected expression of either parent, i.e. $\mu_{g3} > \max\{\mu_{g1},\mu_{g2}\}$ or, equivalently, $\delta_g > |\alpha_g|$. 
% and high-parent  heterosis (HPH) of gene expression occurs when $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$. For simplicity, we define \emph{heterosis} of gene expression to be the event of LPH or HPH. The most important hypotheses are
% \begin{align}
% H_{g0}:&\mu_{g1}\le \mu_{g3}\le \mu_{g2} \mbox{ or } \mu_{g1}\ge \mu_{g3}\ge \mu_{g2} \mbox{\ \ vs.\ \ } \nonumber \\
% H_{g1}:&\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3},
% \label{e:hypotheses}
% \end{align}
% where $H_{g0}$ indicates no heterosis for gene $g$ and $H_{g1}$ indicates  heterosis for gene $g$. 
% We evaluate these hypotheses based on empirical Bayes estimates of their posterior probabilities, e.g., 
% \begin{align}
% P\left(H_{g1}|y, \hat{\pi}_{MAP}\right) &= P\left(\left.\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3}\right| y, \hat{\pi}_{MAP}\right) \nonumber \\
% &\approx \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\mu_{g3}^{(m)}< \min\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) \mbox{ or } \max\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) < \mu_{g3}^{(m)}\right), \label{e:probs}
% \end{align}
% where $\mu_g^{(m)} = \left(\mu_{g1}^{(m)},\mu_{g2}^{(m)},\mu_{g3}^{(m)}\right)$ is the $m^{th}$ MCMC sample from the empirical Bayes posterior, and $\mathrm{I}(A)$ is 1 if A is true and 0 otherwise. 
We evaluate these probabilities based on empirical Bayes estimates of their posterior probabilities, e.g., 
\begin{equation}
%P(LPH_g|y) \approx 
P(\mbox{LPH}_g|y,\hat{\pi},\hat{\gamma}) = 
%P\left(\left. \mu_{g3} < \min\{\mu_{g1},\mu_{g2}\}\right|y,\hat{\pi},\hat{\gamma}\right) = 
P\left(\left.\delta_g < -|\alpha_g|\right|y, \hat{\pi},\hat{\gamma}\right) \approx \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\delta_g^{(m)} < -\left|\alpha_g^{(m)}\right|\right), 
\label{e:probs}
\end{equation}
where $\left(\delta_g^{(m)},\alpha_g^{(m)}\right)$ is the $m^{th}$ MCMC sample from the empirical Bayes posterior, and $\mathrm{I}(A)$ is 1 if A is true and 0 otherwise. HPH probability is defined similarly with the inequality reversed and without the negative sign. We construct a ranked list of genes according to the maximum of the gene's LPH and HPH heterosis probabilities. Geneticists can use this list to prioritize future experiments to understand the molecular genetic mechanisms for heterosis.

@@

Using these parameters and normalization factors, we simulated data according to the negative binomial model in equation \eqref{e:data} independently for each gene. For each simulation, we analyzed a subset of 25,000 genes selected randomly from genes with simulated counts at least one on average and with at most two zeros for each variety across replications.  We repeated this simulation process 10 times for each of 4, 8, and 16 replicates per variety, reusing normalization factors when necessary.

@@

We compared our method to that of \cite{ji2014estimation}, which assumes normality in the response, by modeling the logarithm of the RNAseq count plus one scaled by the normalization factor, i.e. $\log(Y_{gvi}+1)-\gamma_{vi}$. Use of the normalization factor here provides two advantages: 1) counts are properly adjusted for the thoroughness of the sequencing of the sample and 2) for genes with no count variation within variety (which actually occurs in our maize data set), use of the normalization factors allows the approach of \citeauthor{ji2014estimation} to still execute.

@@

With 4 replicates per variety, there does not appear to be much of a difference between \ShrinkBayes{} and the eBayes approaches, but as the number of replicates increases, the eBayes approaches appear to improve relative to \ShrinkBayes{}. Two differences exist between the \ShrinkBayes{} and eBayes approaches that could explain the difference in AUC: 1) \ShrinkBayes{} utilizes a different empirical Bayes approach for estimating both the hyperparameters and the gene-specific parameters and 2) the measure of EH is slightly different due to INLA not providing a full posterior.

There also appears to be a pattern of the eBayes (Laplace) systematically performing better than eBayes (normal). We suspect this is because the Laplace distributions are better approximations to the true underlying distribution for these parameters and we discuss this in Section \ref{s:discussion}.

@@

We used our method to analyze a maize data set \citep{paschold2012complementation} of \RNAseq{} gene expression in parental lines B73 and Mo17 and the hybrid genotype (B73$\times$Mo17) with a total of 39,656 genes. Each variety had four biological replicates measured with Illumina methodology and equipment. Reads were mapped to the whole reference genome using the short reads aligner, NOVOALIGN. For more specifics, please see \cite{paschold2012complementation}.

We analyzed the data using all the methods compared in the previous section. The computation time on a desktop with two 4-core 3.6GHz Intel Xeon processor was 14 seconds for \edgeR{}, 1.3 minutes for \citeauthor{ji2014estimation}, 10 hours for the eBayes approaches, and 17 hours for \baySeq{}. In the eBayes approach, the vast majority of the time is spent on independent MCMC analysis for each gene. Thus we parallelized this step using {\tt doMC} \citep{revolution2014domc} and {\tt plyr} \citep{hadley2011split}. When parallelized across 5 cores, the eBayes approaches took about 2.5 hours. ShrinkBayes took 12 hours on a cluster node with two 8-core 2.6GHz Intel Haswell E5-2640 v3 processors where the code was also parallelized across 5 cores.

For the eBayes (Laplace), we estimated the hyperparameters to be $\hat{\eta}_\phi = 3.44$, $\hat{\sigma}_\phi = 2.74$, $\hat{\eta}_\alpha = -0.05$, $\hat{\sigma}_\alpha=0.36$, $\hat{\eta}_\delta = 0.08$, $\hat{\sigma}_\delta=0.33$, $\hat{\eta}_\psi = -2.35$, and $\hat{\sigma}_\psi=0.60$. We then performed independent MCMC analysis for each gene conditional on these hyperparameters. As with the simulation study, we ran 4 chains simultaneously and doubled the number of MCMC iterations until each gene-specific parameter had an effective sample size above 1,000.

Figure \ref{f:gene_specific_estimates} provides point estimates of the gene-specific parameters from the initial \edgeR{} estimation step and after the eBayes (Laplace) procedure described in Section \ref{s:ebayes}. 
\begin{figure}
\centerline{\includegraphics[width=\textwidth]{gene_specific_estimates.pdf}}
\caption{Two-dimensional histogram of point estimates from \edgeR{} and posterior means from eBayes (Laplace) along with the $y=x$ diagonal.}
\label{f:gene_specific_estimates}
\end{figure}
This figure shows shrinkage for large absolute values of $\alpha$ and $\delta$ from the \edgeR{} estimation toward $\hat{\eta}_\alpha$ and $\hat{\eta}_\delta$ which are both approximately zero. The figure also shows decreased values for the overdispersion parameter with larger decreases for high and low values of overdispersion. Finally, very little, if any, shrinkage is observed for $\phi$.

With posteriors for all parameters, we can calculate empirical Bayes posterior probabilities for LPH and HPH. For each gene, the quantity of interest is likely to be the maximum of these two probabilities. For each gene with a high probability of either HPH or LPH, the magnitude of the effect is of interest, thus we calculate 
\begin{equation}
% \mbox{estimated effect size}_g = \left\{ 
% \begin{array}{ll}
% \hat{\mu}_{g3} - \min(\hat{\mu}_{g1},\hat{\mu}_{g2}) & \mbox{if } \hat{\mu}_{g3} < \min(\hat{\mu}_{g1},\hat{\mu}_{g2}) \\
% \hat{\mu}_{g3} - \max(\hat{\mu}_{g1},\hat{\mu}_{g2}) & \mbox{if } \hat{\mu}_{g3} > \max(\hat{\mu}_{g1},\hat{\mu}_{g2}) \\
% 0 & \mbox{otherwise},
% \end{array} 
% \right. 
\mbox{estimated effect size}_g \equiv \left\{ 
\begin{array}{ll}
\hat{\delta}_g - |\hat{\alpha}_g| & \mbox{if } \hat{\delta}_g > \phantom{-}|\alpha_g| \\
\hat{\delta}_g + |\hat{\alpha}_g| & \mbox{if } \hat{\delta}_g < -|\alpha_g| \\
0 & \mbox{otherwise},
\end{array}.
\right. 
\label{e:effect_size}
\end{equation}
This estimated effect size is the difference between hybrid mean and the nearest parent with negative values indicating LPH and positive values indicating HPH. If the hybrid mean is estimated to be between the parents, this effect is defined to be zero.

@@

Although our method appears to hold some advantage over existing methods, we believe our approach can be improved by refining the hierarchical model for the gene-specific parameter distribution. Figure \ref{f:scatterplot} shows marginal and bivariate histograms for eBayes (Laplace) posterior means for the gene-specific parameters. 
\begin{figure}
\centerline{\includegraphics[width=\textwidth]{estimates}}
\caption{Marginal and bivariate histograms of posterior means for gene-specific parameters for the B73 $\times$ Mo17 maize experiment.}
\label{f:scatterplot}
\end{figure}
These figures show markedly departures from marginal model assumptions, e.g. normality assumptions for $\phi_g$ and $\psi_g$, and independence assumptions for $(\phi_g,\psi_g)$ and $(\alpha_g,\delta_g)$. The plot of $\phi_g$ versus $\psi_g$ shows a pattern where the mean overdispersion decreases as the mean expression level increases. The plot of $\alpha_g$ versus $\delta_g$ shows a rotated V pattern where $\delta_g$ appears to be equal to $|\alpha_g|$. This V pattern is consistent with Mendel's Law of Dominance where the hybrid has mean expression equal to the parent with higher mean expression.

In addition to improving the hierarchical distribution, we believe better estimates of the parameters of this distribution, i.e. the hyperparameters, will also improve detection of gene-expression heterosis. Our current method, based on moment matching of essentially independently estimated gene-specific parameters, provides consistent estimators as the number of replicates per variety increases. But typically the number of replicates per variety is quite small. In our data there are only four replicates per variety, and therefore asymptotic justifications are deficient. There are a variety of alternative estimation approaches to explore, e.g. expectation-maximization algorithms or fully Bayesian approaches.
